GBM Recurrence Proteogenomics

### 2023-02-16 reports/gbm_recurrence_proteogenomics_manuscript_report.qmd
proteogenomics_clean_up

---
title: "GBM Proteomics and Proteolytic processing data analysis - Reproducible report"
author: "Miguel Cosenza"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document
always_allow_html: true
---

```{r}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```

```{r results='hide'}
## Required packages ----
library(tidyverse)
library(mixOmics)
library(fs)
library(kableExtra)
library(sva)
library(limma)
library(naniar)
library(missForest)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(DT)
library(here)
library(janitor)
library(drawProteins)
library(seqinr)
library(ggpubr)

source(here("scr/helper_functions.R"))

theme_set(theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 6, 
                                           angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0, size = 6),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=8),
                legend.text = element_text(size = 7),
                legend.title = element_text(size = 8),
                legend.key.height= unit(3, 'mm'),
                legend.key.width= unit(3, 'mm'),
                legend.position="bottom"))
```

The following report holds the code, results and interpretation of the proteomics data analysis of Initial-Recurrent glioblastoma samples of matched patients.

# Sample annotation info

```{r echo=TRUE}
sample_annotation <- read_csv(here("data/sample_annotation.csv"))

sample_annotation2 <- sample_annotation %>%
  mutate(patient = paste("x",patient, sep = ""),
         recurrence = case_when(recurrence == "initial" ~ "prim",
                                recurrence == "recurrent" ~ "rec",
                                TRUE ~ recurrence)) %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_")) %>%
  filter(recurrence %in% c("prim", "rec"))

## Sample annotation

### Brief note on sample annotation:

Twenty-two patient matched samples (corresponding to 11 patients) were processed for protein extraction, TMT-labelled and submitted via high-pH fractionation and further analyzed via LC-MS/MS.

```{r echo=FALSE, error=TRUE}
kbl(sample_annotation, booktabs = TRUE) %>%
  kable_styling(font_size = 14, bootstrap_options = c("striped", "hover"))
```

# Brief note on data analysis and processing:

Three proteomics data analysis approaches were used to explore the spectral data and will be briefly described:

1. General large scale proteomics.
2. Proteolytic processing.
3. Proteogenomics

The FragPipe bioinformatic pipeline was used for each of these approaches with varying parameters depending on the type of explorative approach.

__1. General large scale proteomics analyses__ : In order to explore the differential abundance of proteins between Recurrent and Initial tumor, a fully-tryptic search was performed against the EBI Human canonical proteome (version 2021_03), appended with common contaminants and iRT peptides. A minimum percentage of  %5 of the total summed reporter ion intensities were required to consider a peptide and reporter ion for quantitation.

__2. Proteolytic processing__: The large scale evaluation of differential proteolytic activity between recurrent vs primary tumor was executed via a semi-tryptic search against the EBI Human canonical proteome (version 2021_03), appended with common contaminants and iRT peptides.  A minimum percentage of  %5 of the total summed reporter ion intensities were required to consider a peptide and reporter ion for quantitation.

__3. Proteogenomics__: The explorative proteogenomics analysis started with the generation of a GBM-specific database from publicly available RNA-seq sequencing data from tumor specimens. Fastq files were obtain from the SRA BioProject PRJNA627121 (Cell Rep, 2021 Mar 2;34(9):108787). Eight samples containing paired reads were selected, based on: 1) the higher number of reads per file; 2) having a similar proportion of males and females and 3) not bearing isocitrate-dehyderogenase (IDH)-1 mutation (IDH WT). None of these patients undergo previous treatment, therefore potential variants in the sequencing data are not expected to be associated to treatment-related mutagenic processes. Spectral data was then searched against this custom database using a fully-tryptic approac. No minimum intensity threshold was set for reporter ion intensity for quantitation.

Exceptionally when stated differently within each sub-approach described above, all these searches were executed using the following parameters: MSFragger was used as a search engine, setting a fully or semi-tryptic specificity allowing for 1 missed cleavages. Precursor mass tolerance was set to -20/20 and fragment mass tolerance of 20 ppm with mass calibration and parameter optimization resulting in 5 ppm. Peptide N-terminal acetylation and peptide N-terminal TMT labeling were set as variable modifications. TMT labelling at K and carbamidomethylation of C were set as fixed modifications. MSBooster was used for deep-learning based predictions of retention time and spectra. Predicted features were used by Percolator for post-processing scoring and false discovery rate (FDR) control via target-decoy competition of all peptide-to-spectrum matches (PSMs) obtained from MSFragger search. ProteinProphet was used as a protein inference algorithm keeping all PSMs with probability score bigger than 0.9. Final report and FDR estimation was based on the filtered PSM and protein lists before protein scoring.

Relative quantitation of identified peptides within each sample was performed via their reported ion intensities using TMT-integrator. Only PSMs coming from unique peptides with a minimum probability of 0.9 and an isotopic purity of at least 50% were considered for quantitation. Quantitative values were normalized via median centering and summarized by protein/gene and peptide using a virtual reference channel, to finally generate protein and peptide normalized abundance matrices of abundance used for further statistical processing in combination with the sample annotation.

# General proteomics analysis

## Initial data loading and wrangling

```{r message=FALSE}
## tmt-integrator output loading 
tmt_protdata_frag <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/tmt-report/abundance_protein_MD.tsv")) %>% 
  janitor::clean_names()

identified_proteins <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_protein.tsv")) %>% 
  janitor::clean_names()

identified_peptides <- read_tsv(here("data/specific_search_fragpipe17/specific_no_ptms_2/combined_peptide.tsv")) %>%
  janitor::clean_names()

fasta_ident_prots <- read.fasta(here("data/specific_search_fragpipe17/specific_no_ptms_2/mix_2/protein.fas"), seqtype = "AA", as.string = TRUE)

@@

Wrangle and merge protein ID/quant with sample annotation

```{r}
# wide expression matrix with NAs
expr_matrix <- dplyr::select(tmt_protdata_frag,
                             Protein = index, starts_with("x"))

long_mat <- pivot_longer(expr_matrix,
                         cols = starts_with("x"),
                         names_sep = "_",
                         names_to = c("patient", "recurrence"),
                         values_to = "Abundance")

long_mat_2 <- long_mat %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_"))

paired_annot_uniq <- long_mat_2 %>%
  dplyr::select(patient, recurrence, paired_id) %>%
  distinct()

sample_annotation2 <- sample_annotation %>%
  mutate(patient = paste("x",patient, sep = ""),
         recurrence = case_when(recurrence == "initial" ~ "prim",
                                recurrence == "recurrent" ~ "rec",
                                TRUE ~ recurrence)) %>%
  mutate(paired_id = paste(patient, recurrence, sep = "_")) %>%
  filter(recurrence %in% c("prim", "rec"))

quant_annotated <- left_join(long_mat_2,
                             sample_annotation2, 
                             by = c("paired_id", "patient","recurrence")) %>%
  mutate(Channel_mix = paste(mixture,channel,  sep = "_"))

## Wide xpression matrices without NAs
expr_matnona <- column_to_rownames(expr_matrix,
                            "Protein") %>%
          as.matrix() %>%
          na.omit()
```

## ID coverage

### Coverage plot

```{r}
ids_summary <- quant_annotated %>% 
  dplyr::select(Protein, mixture, Abundance) %>% 
  distinct(Protein, mixture, .keep_all = TRUE) %>%
  group_by(mixture) %>%
  summarise(`Nr of IDs` = sum(!is.na(Abundance))) %>% 
  bind_rows(.,
            tibble(mixture = "In all three",
                   `Nr of IDs` = nrow(expr_matnona))) %>%
  bind_rows(.,
            tibble(mixture = "Overall",
                   `Nr of IDs` = nrow(expr_matrix)))
```

* Number of identified proteins: `r nrow(expr_matrix)`

* Number of identified and quantified proteins in all samples: `r nrow(expr_matnona)`

The initial analysis with MaxQuant yielded 2090 IDs in all samples (without missing values) and 5176 overall.

```{r fig.height=4, fig.width=7}
ggplot(data = ids_summary,
       aes(y = `Nr of IDs`, x = mixture)) +
  geom_col(fill = "#008683") + 
  geom_text(size = 5, position = position_stack(vjust = 0.5), aes(label = `Nr of IDs`)) +
  coord_cartesian(ylim = c(0, 5000))+
  coord_flip() +
  labs(title = "Number of Identified and quantified Proteins by Mixture") +
  ylab("Nr of IDed and Quant Proteins") + 
  xlab("TMT Mixture") + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 0),
        axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 0),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "grey"),
        panel.border = element_rect(colour = "black", fill=NA, size=1.5),
        axis.title=element_text(size=12,face="bold"))
```  
## Exploratory analysis and visualizations

### Visualization of missing values

```{r fig.height=7, fig.width=8}
vis_miss(expr_matrix) + 
          ggtitle("Distribution of missing values") + 
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))
```

### QC plots after TMT integrator pre-processing and normalization (Protein level)  {.tabset .tabset-fade}

#### By TMT channel/sample {.unnumbered}

##### General Abudance distribution by TMT channel/sample {.unlisted .unnumbered}

```{r fig.height=6, fig.width=9, warning=FALSE}
qcplot <- ggplot(quant_annotated,
                 mapping = aes(x = Channel_mix, 
                               y = Abundance, 
                               fill = recurrence)) +
          geom_boxplot() + 
          #facet_grid(.~recurrence, scales = "free") +
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))

print(qcplot)

The median-centering after TMT-integrator seem to show a slight mixture-dependent centered median per sample. The paired distribution of initial and recurrent tumor allows the biological effect to be more important than any potential TMT-mixture-based batch effect.

The different abundance distribution can also be explained by a different distribution of missing values.

PCA and sPLS-DA analysis allow to confirm this.

#### By Protein (General)

```{r}
# select/sample N proteins to observe their abundance distribution  
which_prots <- sample_proteins(x = quant_annotated$Protein,
                               size = 50,
                               seed = 156)
```

`r length(which_prots)` proteins were randomly sampled to check for the normality of their 'normalized' Abundance distribution.

```{r fig.height=5.5, fig.width=8, warning=FALSE}

qcplot_prots <- ggplot(quant_annotated %>% 
                         filter(Protein %in% which_prots),
                 mapping = aes(x = reorder(Protein,Abundance,na.rm = TRUE), 
                               y = Abundance)) +
          geom_boxplot() + 
          geom_jitter(color="black", size=0.4, alpha=0.9) +
          labs(subtitle = "Protein IDs arranged by median abudance",
               x = "Protein ID ") + 
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))

print(qcplot_prots)
```

#### By Protein and Batch {.unnumbered}

`r length(which_prots)` proteins were randomly sampled to check for the normality of their 'normalized' Abundance distribution.

```{r fig.height=5, fig.width=10, warning=FALSE}
qcplot_prots2 <- ggplot(quant_annotated %>% 
                         filter(Protein %in% which_prots),
                 mapping = aes(x = reorder(Protein,Abundance,na.rm = TRUE), 
                               y = Abundance, 
                               fill = mixture)) +
          geom_boxplot() + 
          geom_jitter(color="black", size=0.4, alpha=0.9) +
          labs(subtitle = "Protein IDs arranged by median abudance",
                       x = "Protein ID ") + 
          facet_grid(.~mixture, scales = "free") +
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))

print(qcplot_prots2)
```

The abundance distribution of individual proteins can be considered as normaly distributed in general, which makes then suitable for inferential statistics via linear models.

### Exploratory PCA analisys

```{r echo=FALSE}
## Get predictors for mixtures
expr_matrix2 <- expr_matrix %>%
  dplyr::select(Protein, sample_annotation2$paired_id)

paired_id <- sample_annotation2$paired_id
recurrence <- sample_annotation2$recurrence
patient <- sample_annotation2$patient
mixture <- sample_annotation2$mixture

## get predictor for mixtures wo patient 6
expr_matrixwo6 <- expr_matrix %>%
  dplyr::select(Protein, sample_annotationwo6$paired_id)

paired_idwo6 <- sample_annotationwo6$paired_id
recurrencewo6 <- sample_annotationwo6$recurrence
patientwo6 <- sample_annotationwo6$patient
mixturewo6 <- sample_annotationwo6$mixture

## Prep matrix without missing values for PCA

expr_matnona <- column_to_rownames(expr_matrix2,
                            "Protein") %>%
          as.matrix() %>%
          na.omit()

colnames(expr_matnona) <- paste(sample_annotation2$paired_id, 
                                sample_annotation2$mixture,
                                sep = "_")

t_expr_matnona <- t(expr_matnona)

expr_mat <- column_to_rownames(expr_matrix2,
                               "Protein") %>%
          as.matrix()

t_expr_mat <- t(expr_mat)

## Matrix without patient 6

expr_matnonawo6 <- column_to_rownames(expr_matrixwo6,
                            "Protein") %>%
          as.matrix() %>%
          na.omit()

colnames(expr_matnonawo6) <- paste(sample_annotationwo6$paired_id, 
                                sample_annotationwo6$mixture,
                                sep = "_")

t_expr_matnonawo6 <- t(expr_matnonawo6)

expr_matwo6 <- column_to_rownames(expr_matrixwo6,
                               "Protein") %>%
          as.matrix()

t_expr_matwo6 <- t(expr_matwo6)
```

#### General PCA

##### Percentage of explained variance per component

The PCA was applied on the expression matrix without missing values

```{r}
pca_res = pca(t_expr_matnona, ncomp = 10, center = TRUE, scale = TRUE)
```

```{r fig.width=7, fig.height=5}
plot(pca_res)
```

~45% of variance based on protein abundance can be explained by components 1 and 2 after PCA.

##### PCA plot on samples based on protein abundance

```{r}
# preprocess pca results 
pca_variates <- pca_res$variates$X %>% 
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  separate(col = "Sample",
           into = c("sample", "recurrence", "mixture"),
           sep = "\\_", extra = "merge") %>%
  mutate(paired_id = paste(sample, recurrence, sep = "_")) %>%
  dplyr::select(-c(mixture, recurrence)) %>%
  left_join(. , sample_annotation2,
            by = "paired_id") %>%
  mutate(Stage = if_else(condition = recurrence == "prim", 
                         true = "Initial", 
                         false = "Recurrent"),
         Mix = case_when(mixture == "tmt_1" ~ "Mix1",
                         mixture == "tmt_2" ~ "Mix2",
                         mixture == "tmt_3" ~ "Mix3"))

pca_variateswopat6 <- pca_variates %>% 
  filter(patient != "x6")

```{r fig.width=7.3, fig.height=5}
ggplot(data = pca_variates,
       aes(x = PC1, y = PC2)) +
  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
  geom_text(aes(label = patient), position = position_nudge(x = 4.5)) +
  labs(title = "PCA plot of samples\nBased on the abundance values of proteins") +
  xlab(paste("PC1", round(pca_res$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("PC2", round(pca_res$prop_expl_var$X[2]*100), "% var explained")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

There is no evident batch effect after PCA analysis.

There is a good number of individual patients that show important differences in their PCA variates values in recurrent vs primary.

```{r fig.width=7.3, fig.height=5}
ggplot(data = pca_variateswopat6,
       aes(x = PC1, y = PC2)) +
  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
  geom_text(aes(label = patient), position = position_nudge(x = 4.5)) +
  labs(title = "PCA plot of samples\nBased on the abundance values of proteins (patient 6 excluded)") +
  xlab(paste("PC1", round(pca_res$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("PC2", round(pca_res$prop_expl_var$X[2]*100), "% var explained")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

```{r}
pca_plot <- ggplot(data = pca_variateswopat6,
       aes(x = PC1, y = PC2)) +
  geom_point(size = 4, aes(shape = Mix, color = Stage)) +  
  geom_text(aes(label = patient), position = position_nudge(x = 4.5), size = 2) +
  xlab(paste("PC1", round(pca_res$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("PC2", round(pca_res$prop_expl_var$X[2]*100), "% var explained")) +
theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 7, angle = 360),
      axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 7),
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.border = element_rect(colour = "black", fill=NA, size=0.5),
      axis.title = element_text(size = 7),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 8),
      legend.key.height= unit(3, 'mm'),
      legend.key.width= unit(3, 'mm'),
      legend.position="bottom")
```

```{r}
ggsave(plot = pca_plot, 
       filename = here::here("figures/pca_proteomics_gbm.tiff"), 
       device = "tiff",
       units = "mm",
       width = 63,
       height = 50)

ggsave(plot = pca_plot, 
       filename = here::here("figures/pca_proteomics_gbm.eps"), 
       device = "eps",
       units = "mm",
       width = 63,
       height = 50)
```

### Exploratory sPLS-DA

```{r}
plsda_rest1 <- plsda(X = t_expr_matnona,
                     Y = recurrence,
                     ncomp = 3)
```

```{r fig.width=6, fig.height=5}
plotIndiv(plsda_rest1,
          nd.names = TRUE, ellipse = TRUE, legend = FALSE, title = "sPLS-DA_Rec-Prim", 
          size.title = rel(2.2), size.xlabel = rel(1.5), size.ylabel = rel(1.5), 
          siz.axis = rel(1.2), point.lwd = 0.8, cex = 4)

Patient 6 in recurrence behaves suspiciously similar to initial samples. We consider this to be an outlier associated to a potential problem during sample prep.

Therefore we decided to exclude patient 6 from further analyses.

```{r}
plsda_rest2 <- plsda(X = t_expr_matnona,
                     Y = mixture,
                     ncomp = 3)
```

```{r fig.width=6, fig.height=5}
plotIndiv(plsda_rest2,
          nd.names = TRUE, ellipse = TRUE, legend = FALSE, title = "sPLS-DA_Rec-Prim", 
          size.title = rel(2.2), size.xlabel = rel(1.5), size.ylabel = rel(1.5), 
          siz.axis = rel(1.2), point.lwd = 0.8, cex = 4)

We can say that the biological differences between initial and recurrent status are stronger than the differences that could come from potential batch effects.

### Exploratory sPLS-DA without patient 6

```{r}
plsda_restwo6 <- plsda(X = t_expr_matnonawo6,
                     Y = recurrencewo6,
                     ncomp = 3)
```

```{r}
plotIndiv(plsda_restwo6,
          nd.names = TRUE, ellipse = TRUE, legend = FALSE, title = "sPLS-DA_Rec-Prim wo Pat6", 
          size.title = rel(2.2), size.xlabel = rel(1.5), size.ylabel = rel(1.5), 
          siz.axis = rel(1.2), point.lwd = 0.8, cex = 4)

#### GGplot-based visualization of Exploratory sPLS-DA

```{r}
# preprocess pca results 
plsda_variates <- plsda_restwo6$variates$X %>% 
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  separate(col = "Sample",
           into = c("sample", "recurrence", "mixture"),
           sep = "\\_", extra = "merge") %>%
  mutate(paired_id = paste(sample, recurrence, sep = "_")) %>%
  dplyr::select(-c(mixture, recurrence)) %>%
  left_join(. , sample_annotation2,
            by = "paired_id")

plsda_variateswpat6 <- plsda_rest1$variates$X %>% 
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  separate(col = "Sample",
           into = c("sample", "recurrence", "mixture"),
           sep = "\\_", extra = "merge") %>%
  mutate(paired_id = paste(sample, recurrence, sep = "_")) %>%
  dplyr::select(-c(mixture, recurrence)) %>%
  left_join(. , sample_annotation2,
            by = "paired_id")

```{r fig.width=7.3, fig.height=5}
ggplot(data = plsda_variateswpat6,
       aes(x = comp1, y = comp2)) +
  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
  geom_text(aes(label = patient), position = position_nudge(x = 2)) +
  labs(title = "sPLS-DA plot of samples\nBased on the abundance values of proteins") +
  xlab(paste("sPLS-DA Comp-1", round(plsda_restwo6$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("sPLS-DA Comp-2", round(plsda_restwo6$prop_expl_var$X[2]*100), "% var explained")) + 
  ggforce::geom_mark_ellipse(mapping = aes(color = recurrence)) +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

Patient 6 in recurrence behaves suspiciously similar to initial samples. We consider this to be an outlier associated to a potential problem during sample prep.

Therefore we decided to exclude patient 6 from further analyses.

```{r fig.width=7.3, fig.height=5}
ggplot(data = plsda_variates,
       aes(x = comp1, y = comp2)) +
  geom_point(size = 4, aes(shape = mixture, color = recurrence)) +  
  geom_text(aes(label = patient), position = position_nudge(x = 2)) +
  labs(title = "sPLS-DA plot of samples - pat6 excluded \nBased on the abundance values of proteins") +
  xlab(paste("sPLS-DA Comp-1", round(plsda_restwo6$prop_expl_var$X[1]*100), "% var explained")) + 
  ylab(paste("sPLS-DA Comp-2", round(plsda_restwo6$prop_expl_var$X[2]*100), "% var explained")) + 
  ggforce::geom_mark_ellipse(mapping = aes(color = recurrence)) +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom")
```

The difference between initial and recurrent tumor is better resolved in the sPLS-DA plot when excluding patient 6.

## Sparcity reduction

Only proteins that are identified and quantified in 2 out of 3 TMT mixtures will be kept for the inferential analysis. The proteins missing in 1 mixture will be imputed with `missForest`.

__Function to select proteins to exclude:__

```{r}
sel_proteins_missing <- function(long_matrix,
                                 threshold) {

na_count <- group_by(long_matrix,
                     Protein, mixture) %>%
    summarise(na_count = sum(is.na(Abundance)),
              total = n()) %>% 
    ungroup() %>% 
    mutate(NA_fraction = na_count/total)

na_count_perbatch <- na_count %>%
  group_by(Protein) %>%
  summarise(na_per_batch = sum(NA_fraction)) %>%
  ungroup()

proteins2exclude <- na_count_perbatch %>%
  filter(na_per_batch > threshold) %>%
  pull(Protein)

return(proteins2exclude)
}

__Excluded proteins:__

```{r}
proteins2exclude <- sel_proteins_missing(quant_annotated,
                                         threshold = 1)
```

Exclude proteins from initial abundance matrix.

```{r}
expr_matrix_filt <- expr_matrix2 %>%
  filter(!Protein %in% proteins2exclude)
```

### Visualize missing values (before and after filtering)

```{r}
nofiltnas_plot <- vis_miss(expr_matrix) + 
          ggtitle("Distribution of missing values",
                  subtitle = paste("All quant proteins. Total =", nrow(expr_matrix))) + 
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))

filtnas_plot <- vis_miss(expr_matrix_filt) + 
          ggtitle("Distribution of missing values",
                  subtitle = paste("Only present in 2/3 TMT mixtures. Total =", nrow(expr_matrix_filt))) + 
          theme(axis.text.x = element_text(hjust = 0.5, vjust = 0.1, size = 10, angle = 90),
                axis.text.y = element_text(hjust = 0.5, vjust = 0.1, size = 10),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", fill=NA, size=1.5),
                axis.title=element_text(size=12,face="bold"))
```

```{r fig.height=6, fig.width=12}
cowplot::plot_grid(nofiltnas_plot, filtnas_plot, ncol = 2)
```

## `missForest` imputation:

```{r}
mat_filt <- expr_matrix_filt %>%
  column_to_rownames("Protein") %>%
  as.matrix()

t_mat_filt <- t(mat_filt)

if(!file.exists(here("rds/missforest_imp_mat_filt.Rds"))){

 t_imp_mat_filt <- missForest::missForest(t_mat_filt)
 t_imp_filt_mat <- t_imp_mat_filt$ximp

 mat_filt_imp <- t(t_imp_filt_mat)

 saveRDS(t_imp_mat_filt,
         file = here("rds/missforest_imp_mat_filt.Rds"))

} else {

 t_imp_mat_filt <- readRDS(here("rds/missforest_imp_mat_filt.Rds"))
 t_imp_filt_mat <- t_imp_mat_filt$ximp

 mat_filt_imp <- t(t_imp_filt_mat)

}

## Inferential analysis

__Protein to Gene annotation__

```{r}
prot2gene <- tmt_protdata_frag %>%
  dplyr::select(Protein = index, Gene = gene)
```

__Prep protein abundance matrix__

```{r}
mat_filt_impwo6 <- as.data.frame(mat_filt_imp) %>%
  rownames_to_column("Protein") %>%
  dplyr::select(-starts_with("x6")) %>%
  column_to_rownames("Protein") %>%
  as.matrix()

### Set up design matrix

Initial vs recurrent samples are patient-matched, therefore the design matrix is setup so the linear model can account for the patient random effect.

```{r}
design_wo6 <- model.matrix(~patientwo6+recurrencewo6)

rownames(design_wo6) <- sample_annotationwo6$paired_id
```

Check design matrix:

```{r}
print(design_wo6)
```

```{r}
mat_filt_impwo6[1:3,1:12]
```

### Fit limma

```{r}
source(here("scr/fit_limmawo6.R"))
```

```{r}
# fit limma
limma_tab_wo6 <- fit_limmawo6(mat_filt_impwo6, 
                         design_wo6, 
                         method = 'robust', 
                         Limma = "Robust - w Patient effect wo Pat6",
                         prot2gene = prot2gene)

#get proteins increased in recurrence
increased_in_rec <- limma_tab_wo6 %>%
  filter(logFC > 0,
         adj.P.Val < 0.05) %>%
  pull(Protein)

#get proteins decreased in recurrence
decreased_in_rec <- limma_tab_wo6 %>%
  filter(logFC < 0,
         adj.P.Val < 0.05) %>%
  pull(Protein)
```

### Volcano plots

```{r fig.width=8, fig.height=6}
our_volcano(limma_tab_wo6, 
            FC_cutoff = 0, 
            pval_cutoff = 0.05, 
            color_diffex = "red", 
            color_nondifex = "#2a9d8f", 
            interesting_proteins = NULL, 
            vert_line_col = "red",
            hline_col = "red", 
            hline_pos = 0.05, 
            linetype = "dashed",
            increased_in = "Recurrent", 
            comparison_title = "Recurrent vs Initial") +
  theme(axis.text.x = element_text(hjust = 0.5, vjust = 0, size = 10, angle = 360),
        axis.text.y = element_text(hjust = 0.95, vjust = 0.2, size = 10),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),
        legend.position="bottom") 
```

```{r fig.width=8, fig.height=6, echo=FALSE}
protlabels <- limma_tab_wo6 %>%
      filter(logFC > 0 | logFC < 0,
             adj.P.Val < 0.05)

signi_hits <- nrow(protlabels)

signi_up <- limma_tab_wo6 %>%
      filter(logFC > 0 ,
             adj.P.Val < 0.05) %>%
  nrow()

signi_down <- limma_tab_wo6 %>%
      filter(logFC < 0 ,
             adj.P.Val < 0.05) %>%
  nrow()

size <- 0.4

>>>>

---
title: "GBM Recurrence Proteogenomics"
author: "Miguel Cosenza-Contreras"
format:
  gfm:
    toc: true 
    toc-depth: 4 # default is 3
    number-sections: true
    execute:
      output: true
always_allow_html: true
editor: source
---

```{r}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

source(here::here("scr/helper_functions.R"))

## Required packages ----
library(tidyverse)
library(mixOmics)
library(fs)
library(kableExtra)
library(sva)
library(limma)
library(naniar)
library(missForest)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(DT)
library(here)
library(janitor)
library(drawProteins)
library(seqinr)
library(ggpubr)
library(ggrepel)
library(extrafont)
extrafont::loadfonts(device = "win")

theme_set(theme(axis.text.x = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6, 
                                           angle = 360),
                axis.text.y = element_text(hjust = 0.5, 
                                           vjust = 0, 
                                           size = 6),
                panel.background = element_blank(),
                panel.grid.major = element_line(color = "grey"),
                panel.border = element_rect(colour = "black", 
                                            fill = NA, 
                                            size = 1.5),
                axis.title=element_text(size = 8),
                legend.title = element_text(size = 8),
                legend.key.height = unit(3, 
                                         'mm'),
                legend.key.width = unit(3, 
                                        'mm'),
                legend.position = "bottom",
                text = element_text(family = "Helvetica")))

@@

# Initial data loading and wrangling
